---
title: "Fabi's Week 6 Homework: Nutrient Dynamics at Maunalua Bay"
author: "Fabiola Gonzalez Zamora"
format:
  html:
    theme: yeti
    toc: true
    toc-location: left
    embed-resources: true
    code-fold: true
    code-summary: "Show Code"
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.path: "../Output/"
    fig.width: 9
    fig.height: 6
    dpi: 300
---

```{r Load Packages}
#| include: FALSE

library(tidyverse)
library(dplyr)
library(here)
library(beepr)
library(peRReo)
library(kableExtra)
library(cowsay)
```

## Introduction

- Coastal groundwater can transport nutrients to nearshore ecosystems
- This analysis examines **nutrient concentrations** at site BP in Maunalua Bay
  - Comparison of samples during *different tides and times of day* focusing on:
    - Phosphate,
    - Silicate,
    - Nitrate + Nitrite

## Data Preparation

Steps before analysis:

1.  Remove observations with missing values.
2.  Separate `Tide_time` into `Tide` and `Time`.
3.  Retain observations from Site BP.
4.  Reshape three nutrient columns into a tidy, long format.

```{r Load and Clean Data}

# Load csv file
ChemData <- read.csv(here("Week_06", "Data", "chemicaldata.csv"))

# Clean data
ChemData_clean <- ChemData |>
  drop_na() |>       
  separate_wider_delim(cols = Tide_time, 
                       delim = "_",  
                       names = c("Tide", "Time"),
                       cols_remove = FALSE) |>
  filter(Site == "BP") |> 
  pivot_longer(cols = c(Phosphate, Silicate, NN), 
               names_to = "Nutrient", 
               values_to = "Concentration")|>
  mutate( # Replace abbreviated variable names with reader-friendly labels
    Nutrient = recode(
      Nutrient,
      Phosphate = "Phosphate",
      Silicate = "Silicate",
      NN = "Nitrate + nitrite"
    ),
    # Control the order shown in tables and figures
    Tide = factor(Tide, levels = c("Low", "High")),
    Time = factor(Time, levels = c("Day", "Night"))
  )
```

## Summary statistics

- Summary of mean concentration, standard deviation, sample size, and standard error for every nutrient, tide, and time-of-day combination
- Results are presented in @tbl-nutrient-summary

```{r Summary Statistics}
ChemSummary <- ChemData_clean |>
  group_by(Nutrient, Tide, Time) |>
  summarise(mean_concentration = mean(Concentration), 
            var_concentration = var(Concentration),   
            sd_concentration = sd(Concentration),    
            sample_size = n(),
            sd_error = sd_concentration / sqrt(sample_size),
            .groups = "drop") 

# Save the summary statistics in the Week 6 output folder
write_csv(
  ChemSummary,
  here("Week_06", "Output", "chemistry_summary.csv"))
```

```{r tbl-nutrient-summary}
#| tbl-cap: "Nutrient concentrations at Site BP by Tide and Time of Day"
#| echo: false

# Format the summary statistics as a polished HTML table
ChemSummary |>
  mutate(
    mean_concentration = round(mean_concentration, 2),
    sd_concentration = round(sd_concentration, 2),
    sd_error = round(sd_error, 2)
  ) |>
  rename(
    "Nutrient" = Nutrient,
    "Tide" = Tide,
    "Time of day" = Time,
    "Mean" = mean_concentration,
    "Variance" = var_concentration,
    "Standard deviation" = sd_concentration,
    "Sample size" = sample_size,
    "Standard error" = sd_error
  ) |>
  kbl(
    align = "lcccccc",
    booktabs = TRUE
  ) |>
  kable_classic(
    full_width = FALSE,
    html_font = "Arial"
  ) |>
  row_spec(0, bold = TRUE, color = "white", background = "#284C46") |>
  column_spec(1, bold = TRUE)
```

## Nutrient Patterns

- @fig-nutrient-concentrations compares mean nutrient concentrations between high and low tides
  - Bars separated by time of day, error bars represent **one standard error above and below the means**

```{r fig-nutrient-concentrations}
#| fig-cap: "Mean nutrient concentrations at Site BP by tide and time of day. Error bars represent ±1 standard error."
#| echo: false
#| fig-align: center

# Select two contrasting, colorblind-friendly colors
time_colors <- latin_palette("shakira", 2)
names(time_colors) <- c("Day", "Night")

# Create a publication-quality nutrient figure
ChemPlot <- ChemSummary |>
  ggplot(
    aes(
      x = Tide,
      y = mean_concentration,
      fill = Time)) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7,
    alpha = 0.85) +
  geom_errorbar(
    aes(
      ymin = pmax(0, mean_concentration - sd_error),
      ymax = mean_concentration + sd_error,
      color = Time),
    position = position_dodge(width = 0.8),
    width = 0.2,
    linewidth = 0.8,
    show.legend = FALSE) +
  facet_wrap(
    ~ Nutrient,
    scales = "free_y") +
  scale_fill_manual(values = time_colors) +
  scale_color_manual(values = time_colors) +
  labs(
    title = "Nutrient Concentrations Vary with Tide and Time of Day",
    subtitle = "Site BP, Maunalua Bay",
    x = "Tide",
    y = expression("Mean concentration (" * mu * "mol/L)"),
    fill = "Time of day",
    caption = "Data source: Maunalua Bay chemistry dataset") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(
      size = 17,
      face = "bold",
      color = "#244B55"),
    plot.subtitle = element_text(
      size = 12,
      color = "#4A4A4A"),
    axis.title = element_text(
      face = "bold"),
    strip.background = element_rect(
      fill = "#DDECEF",
      color = NA),
    strip.text = element_text(
      face = "bold",
      color = "#244B55"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    panel.grid.major.y = element_line(
      color = "grey90",
      linewidth = 0.3),
    plot.caption = element_text(
      size = 8,
      color = "grey40"))

ChemPlot
```

## Interpretation

The results show that nutrient concentrations are **not uniform** across sampling conditions. Silicate occurs at a much higher concentration than the other nutrients, which is why each panel in @fig-nutrient-concentrations uses its own vertical scale.

Some considerations: - the bars show group means, not individual observations; - error bars show uncertainty around each mean; - this visualization describes patterns but does not test statistical significance.

Overall, the table and figure provide *complementary views* of the same chemistry data: @tbl-nutrient-summary provides precise values, while @fig-nutrient-concentrations makes the major patterns easier to compare.

## Data Sources

The analysis uses the Maunalua Bay chemistry dataset supplied for OCN 682/MBIO 612.
